Aquifer depletion exacerbates agricultural drought losses in the US High Plains

### 2023-10-09 Writing/NatureWater/3rdRoundRevision/replies/reply_to_reviewers_R3.rmd

+ Line 289 – "thickness" should be "thicknesses".

\textcolor{blue}{Thank you for catching this. In the sentence, "thickness" was replaced with "thicknesses".}

+ Figure 7 – need a 200 m label for scale bar on soybean map. Are the blank areas in the aquifer in Texas and Oklahoma an indication that the aquifer is thin or that there is a lack of data? Please clarify.

\textcolor{blue}{200m lable was added to the legend of the soybean map. The blank area are counties that were not used at all in either of the yield and irrigation share regressions because neither of them are reported by USDA. 
}

@@

\textcolor{blue}{Thank you for this suggestion. We modify the sentence as follows:}

\textcolor{blue}{As drought severity increases (i.e., as water deficit increases), rainfed yields of both crops decline significantly whereas irrigated yields are increasing or stable as deficit increases (see Figures A.1 and A.2 for the 95\% confidence intervals)}

+ (2) The beginning of the response letter describes a code and data correction unrelated to the previous reviewer comments. "[Counties] with only irrigated acres" were removed from the irrigation share regression. The immediate acknowledgement of this correction is appreciated. I trust that it was based on sensible considerations, though the rationale does not immediately become clear from the short statement, especially given that this does not seem to have been relevant previously. As you point out, the result changes due to this correction are quantitatively relatively minor (most visible in Figs. 3 and 4), and do not affect the qualitative findings or conclusions.

\textcolor{blue}{We apologize for the confusion we caused. Our sentnce on this issue was not clear in the last round of revision. We meant that we excluded counties where only the irrigated acres are "observed" because only irrigated acres were reported by USDA, but not rainfed acres. It is very much possible that those counties had rainfed production, but rainfed acres were simply not reported by USDA for the counties we dropped. This means that we cannot even calculate the share of irrigation for those counties (This is also the case for counties with only rainfed production observations). So, we were not trying to drop the fully irrigated counties. Of course, they may be fully irrigated and actually had zero rainfed production though it is very unlikely. Previously before fixing the coding error, those counties were assigned the value of 1 (100\%) as their share of irrigated acres just because drayland acres were not repoted by USDA, which can be a significant over-estimation of their actual share of irrigated acres. Therefore, we should drop them to avoid the problem lowering the quality of the data and consequently analysis. This is more of a data limitation than anything.}

@@

\textcolor{blue}{However, we were actually using 12m as the lower bound, which we realized after seeing reviewer's comment on this in the first round:}

\textcolor{blue}{"Figure 2 - Why do the plots stop at a thickness of about 12 m when you had earlier defined the minimum thickness as 9 m and mention 10 m in the text when discussing the relationships?"}

\textcolor{blue}{Therefore, we change the code so that 9m is used instead of 12m. We, however, did not update the map according to this change in the data. This is why the map in this version has several more counties in Kansas and New Mexico that did not appear in the map of the first round revision. They used to be dropped prior to chaging the threshold from 12m to 9m.}

2. \textcolor{blue}{The significant change in the map in the second round from the 1st round is a mistake on using the map we created when we were experimenting with codes. We then did not rerun the code to create a map with the whole dataset to overwrite the wrong map with the correct map. In this round, we made sure that we streamlined all the codes, rerun all of them, and cofirmed that the map indeed reflect what we used for our analysis in this round. Thank you again for your keen eyes on details.}

### 2023-10-09 Writing/NatureWater/3rdRoundRevision/replies/reply_to_reviewers_R3.tex

\textcolor{blue}{Thank you for catching this. In the sentence, "of the aquifer" was removed to avoid redundancy.}

\begin{itemize}
\tightlist
\item
  Line 289 -- ``thickness'' should be ``thicknesses''.
\end{itemize}

\textcolor{blue}{Thank you for catching this. In the sentence, "thickness" was replaced with "thicknesses".}

\begin{itemize}
\tightlist
\item
  Figure 7 -- need a 200 m label for scale bar on soybean map. Are the
  blank areas in the aquifer in Texas and Oklahoma an indication that
  the aquifer is thin or that there is a lack of data? Please clarify.
\end{itemize}

\textcolor{blue}{200m lable was added to the legend of the soybean map. The blank area are counties that were not used at all in either of the yield and irrigation share regressions because neither of them are reported by USDA. 
}

\begin{itemize}
\tightlist
\item
  Line 308 -- ``y'' should be ``Y''.
\end{itemize}

@@

\textcolor{blue}{However, we were actually using 12m as the lower bound, which we realized after seeing reviewer's comment on this in the first round:}

\textcolor{blue}{"Figure 2 - Why do the plots stop at a thickness of about 12 m when you had earlier defined the minimum thickness as 9 m and mention 10 m in the text when discussing the relationships?"}

\textcolor{blue}{Therefore, we change the code so that 9m is used instead of 12m. We, however, did not update the map according to this change in the data. This is why the map in this version has several more counties in Kansas and New Mexico that did not appear in the map of the first round revision. They used to be dropped prior to chaging the threshold from 12m to 9m.}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  \textcolor{blue}{The significant change in the map in the second round from the 1st round is a mistake on using the map we created when we were experimenting with codes. We then did not rerun the code to create a map with the whole dataset to overwrite the wrong map with the correct map. In this round, we made sure that we streamlined all the codes, rerun all of them, and cofirmed that the map indeed reflect what we used for our analysis in this round. Thank you again for your keen eyes on details.}
\end{enumerate}

### 2023-10-09 Writing/NatureWater/3rdRoundRevision/manuscript.tex
final edits by TM before Tim reviews

As a proxy for the extent of aquifer depletion we use annual estimates of aquifer thickness generated by existing research \citep{haacker2016water,haacker2023}. These data provide annual estimates of the aquifer thickness of the aquifer at 250m resolution since 1935. Annual aquifer thickness maps were derived based on interpolation of 14,000 to 20,000 individual observations of water levels per year, along with detailed maps of both the land surface elevation and bedrock elevation of the HPA. Aquifer thickness estimates for each year were aggregated to the county level using an area-weighted average to match the resolution of agricultural production data. Aquifer thickness values were only estimated for counties with irrigated yield observations. Furthermore, during the aggregation process, any aquifer cells within a county that had a aquifer thickness value lower than 9 m were removed before aggregation to reflect the fact that aquifer thickness lower than this threshold are not considered to be viable for extraction and therefore it is unlikely these areas contribute groundwater for irrigation production \citep{fenichel2016measuring, haacker2016water, deines2020transitions}. Figure \ref{fig:sat-map} presents the distribution of aquifer thickness across our study area for counties included in corn and soybean analyses. Aquifer thickness levels vary substantially over space and time, which we will exploit in our analysis to quantify how aquifer thickness influences farmers' exposure and responses to drought.

@@

All the codes used for this study are available at the following Github repository (\href{https://github.com/tmieno2/Drought-Production-Risk-Aquifer}{link}). All analyses were performed in R \citep{R} and all the raw data-sets used in this study are publicly accessible. The boundary shape file of the High Plains aquifer was obtained from the U.S. Geological Survey at \url{https://water.usgs.gov/GIS/metadata/usgswrd/XML/ds543.xml#stdorder}. This data is used for Figure \ref{fig:sat-map}. Corn and soybean yield data were obtained from the U.S. Department of Agriculture's National Agricultural Statistics Service (USDA-NASS), available at \url{https://www.nass.usda.gov/Quick_Stats/}. Aquifer thickness data were obtained from \url{https://www.hydroshare.org/resource/7d925c7944244032af98c9ed20c22db6/} \citep{haacker2023}. The R tidyUSDA package \citep{RtidyUSDA} was used to automatically download the relevant data. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. Weather data (precipitation, temperature, and evapotranspiration) was obtained from the gridMET dataset \citep{Abatzoglou2013}, which is available at \url{https://www.climatologylab.org/gridmet.html}. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. The aquifer thickness data were obtained from Hydroshare \citep{haacker2023} available at \url{https://www.hydroshare.org/resource/7d925c7944244032af98c9ed20c22db6/}. This data is used for Figure \ref{fig:sat-map} and the regression analysis. Soil characteristics data was obtained from the Soil Survey Geographic Dataset, which is used for regression analysis. The R soilDB package \citep{Rsoildb} was used to download the data. Regression analyses are implemented using the R fixest package \citep{Rfixest}. Other key R packages include ggplot2 \citep{Rggplot2} for creating plots, data.table \citep{Rdatatable} for data wrangling, and sf \citep{Rsf} for spatial data operations. The data used for the regression analysis can be generated using our codes by following the instructions available at the linked Github repository.

>>>>

As a proxy for the extent of aquifer depletion we use annual estimates of aquifer thickness generated by existing research \citep{haacker2016water,haacker2023}. These data provide annual estimates of the aquifer thickness at 250m resolution since 1935. Annual aquifer thickness maps were derived based on interpolation of 14,000 to 20,000 individual observations of water levels per year, along with detailed maps of both the land surface elevation and bedrock elevation of the HPA. Aquifer thickness estimates for each year were aggregated to the county level using an area-weighted average to match the resolution of agricultural production data. Aquifer thickness values were only estimated for counties with irrigated yield observations. Furthermore, during the aggregation process, any aquifer cells within a county that had a aquifer thickness value lower than 9 m were removed before aggregation to reflect the fact that aquifer thicknesses lower than this threshold are not considered to be viable for extraction and therefore it is unlikely these areas contribute groundwater for irrigation production \citep{fenichel2016measuring, haacker2016water, deines2020transitions}. Figure \ref{fig:sat-map} presents the distribution of aquifer thickness across our study area for counties included in corn and soybean analyses. Aquifer thickness levels vary substantially over space and time, which we will exploit in our analysis to quantify how aquifer thickness influences farmers' exposure and responses to drought.

@@

All the codes used for this study are available at the following Github repository (\href{https://github.com/tmieno2/Drought-Production-Risk-Aquifer}{link}). All analyses were performed in R \citep{R} and all the raw data-sets used in this study are publicly accessible. The boundary shape file of the High Plains aquifer was obtained from the U.S. Geological Survey at \url{https://water.usgs.gov/GIS/metadata/usgswrd/XML/ds543.xml#stdorder}. This data is used for Figure \ref{fig:sat-map}. Corn and soybean yield data were obtained from the U.S. Department of Agriculture's National Agricultural Statistics Service (USDA-NASS), available at \url{https://www.nass.usda.gov/Quick_Stats/}. The R tidyUSDA package \citep{RtidyUSDA} was used to automatically download the relevant data. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. Weather data (precipitation, temperature, and evapotranspiration) was obtained from the gridMET dataset \citep{Abatzoglou2013}, which is available at \url{https://www.climatologylab.org/gridmet.html}. This data is used for Figure \ref{fig:deficit-yield-hist} and the regression analysis. The aquifer thickness data were obtained from Hydroshare \citep{haacker2023} available at \url{https://www.hydroshare.org/resource/7d925c7944244032af98c9ed20c22db6/}. This data is used for Figure \ref{fig:sat-map} and the regression analysis. Soil characteristics data was obtained from the Soil Survey Geographic Dataset, which is used for regression analysis. The R soilDB package \citep{Rsoildb} was used to download the data. Regression analyses are implemented using the R fixest package \citep{Rfixest}. Other key R packages include ggplot2 \citep{Rggplot2} for creating plots, data.table \citep{Rdatatable} for data wrangling, and sf \citep{Rsf} for spatial data operations. The data used for the regression analysis can be generated using our codes by following the instructions available at the linked Github repository.
